---
title: 'GWAS tutorial: SNP testing (joint approach)'
author: Patrick Breheny and Tabitha Peter 
date: '`r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  chunk_output_type: console
---

```{r knitr_setup, include=FALSE, purl=FALSE}
library(knitr)
library(kableExtra)
set.seed(1)
knitr::opts_knit$set(aliases=c(h = 'fig.height', w = 'fig.width'))
knitr::opts_chunk$set(comment="#", message=FALSE, collapse=TRUE, cache=FALSE, tidy=FALSE, fig.align="center")
knitr::knit_hooks$set(small.mar = function(before, options, envir) {
  if (before) par(mar = c(4, 4, .1, .1))
})
```

The objective of this module is to test for significant SNPs using a joint approach. For this, we will use the `penalizedLMM` R package, which is available [on GitHub](https://github.com/areisett/penalizedLMM). Once again, we will begin by loading the necessary libraries: 

```{r libs, message=FALSE}
library(data.table)
library(magrittr)
library(qqman)
library(snpStats)
library(dplyr)

# devtools::install_github("areisett/penalizedLMM")
library(penalizedLMM)
```

Next, we load the data. Start by loading the clinical data, as this has the outcome (coronary artery disease ('CAD')) we need for our models. 

```{r load_clinical_dat}
clinical <- fread("data/penncath.csv")
# str(clinical) # if you need to remind yourself of what is here
```

We also need to load the genetic data. For this section, we will work with the quality controlled (QC'd) data from the `SNPRelate` package (see module 1). We will also need the ".bim" file from the original data for making plots. Finally, we need our design matrix $X$ with no missing values (this is the $X$ we obtained by our imputation procedures). 
  
```{r load_SNP_dat}
# load QC'd data:
qc_dat <- readRDS('data/gwas-qc.rds')
# load the bim file
bim <- fread('data/penncath.bim')
# load design matrix of SNP data (you would need to complete the imputation module first)
X <- readRDS(file = "data/fully_imputed_numeric.rds")
```



If you completed the population structure module, you should load the principal components as well - we need them for our analysis. If you did not work through the population structure module, you can skip this step.  

```{r load_PC_dat}
# load principal components 
PCs <- readRDS(file = "data/PCs_base.rds") %>%
  as.data.frame()

names(PCs) <- paste0("PC", 1:ncol(PCs))

```


# Constructing the model 

```{r plmm, eval=FALSE}
# the CAD outcome is binary (0/1), so it may not be the best place to begin for the 
#   joint model example 
joint_model <- plmm(X = X,
                    y = clinical$CAD,
                    intercept = FALSE)

```



# Examining the results 

# Comparing the results of the marginal and joint approaches




